***We are Mergin the whole trip data into one big trip data over here from year 2015 - 2022***

```{r}
rm(tripData)
rows = c("tripduration", "starttime", "stoptime", "start.station.id", "start.station.name", "start.station.latitude", "start.station.longitude", "end.station.id", "end.station.name", "end.station.latitude", "end.station.longitude", "bikeid", "usertype")
tripData = data.frame()
tripData = rbind(tripData, read.csv("TripData/2015-tripdata.csv")[rows])
tripData = rbind(tripData, read.csv("TripData/2016-tripdata.csv")[rows])
tripData = rbind(tripData, read.csv("TripData/2017-tripdata.csv")[rows])
tripData = rbind(tripData, read.csv("TripData/2018-tripdata.csv")[rows])
tripData = rbind(tripData, read.csv("TripData/2019-tripdata.csv")[rows])
tripData = rbind(tripData, read.csv("TripData/2020-tripdata.csv")[rows])
tripData = rbind(tripData, read.csv("TripData/2021-tripdata.csv")[rows])
tripData = rbind(tripData, read.csv("TripData/2022-tripdata.csv")[rows])
head(tripData)
```

***Writing all data inside one huge file***

```{r}
write.csv(tripData, "TripData/tripdata.csv")
```

*Read data if already got written for Trip Data.*

```{r}
rm(tripData)
tripData = read.csv("TripData/tripData.csv")
tripData
cat("Number of na values inside the data", sum(is.na(tripData)))
```

***Counting number of trips by date***

```{r}
library(dplyr)
library(ggplot2)
tripData$date = as.Date(tripData$starttime, "%Y-%m-%d")

numberOfTrips = tripData %>% group_by(date) %>% 
  summarise(totalTrips=length(bikeid),
            .groups = 'drop')
numberOfTrips
ggp = ggplot(data = numberOfTrips) + 
  geom_line(aes(x = date, y = totalTrips), color = 2) + 
  xlab("Year") + ylab("Total Number of Trips") + ggtitle("Date Vs Number of Trips")
ggp
```

```{r}
crimeData = read.csv("CrimeData/crimeData.csv")
head(crimeData)
crimeData
```

***Getting the Crime Count occured on a date***
```{r}
library(dplyr)
library(ggplot2)
crimeData$OCCURRED_ON_DATE = as.Date(crimeData$OCCURRED_ON_DATE, "%Y-%m-%d")
crimeCount = crimeData %>% group_by(OCCURRED_ON_DATE) %>% 
  summarise(totalCrimes=length(INCIDENT_NUMBER),
            .groups = 'drop')
crimeData
crimeCount = crimeCount[crimeCount$OCCURRED_ON_DATE >= as.Date("2015-01-01") & crimeCount$OCCURRED_ON_DATE <= as.Date("2022-12-31"), ]
length(crimeCount$totalCrimes)
length(crimeCount$OCCURRED_ON_DATE)
# plot(x = crimeCount$OCCURRED_ON_DATE, y = crimeCount$totalCrimes, xlab = "Month of the Year", ylab = "Total Number of Crimes", type = "l", main = "Date Vs Number of Crimes")
ggp = ggplot(data = crimeCount) +
  geom_line(aes(x = OCCURRED_ON_DATE, y = totalCrimes), color = 4) +
  xlab("Year") + ylab("Tatal Number of Crimes")
ggp
```

***Combining the weather data***
```{r}
rm(weatherData)
weatherData = data.frame()
weatherData = rbind(weatherData, read.csv("WeatherData/Weather2015.csv"))
weatherData = rbind(weatherData, read.csv("WeatherData/Weather2016.csv"))
weatherData = rbind(weatherData, read.csv("WeatherData/Weather2017.csv"))
weatherData = rbind(weatherData, read.csv("WeatherData/Weather2018.csv"))
weatherData = rbind(weatherData, read.csv("WeatherData/Weather2019.csv"))
weatherData = rbind(weatherData, read.csv("WeatherData/Weather2020.csv"))
weatherData = rbind(weatherData, read.csv("WeatherData/Weather2021.csv"))
weatherData = rbind(weatherData, read.csv("WeatherData/Weather2022.csv"))

weatherData$datetime = as.Date(weatherData$datetime, "%Y-%m-%d")
head(weatherData)
```


***Write the merged data into one big file. Now you can just get if from over here***
```{r}
write.csv(weatherData, "WeatherData/weather.csv")
```

***Read CSV if got already written.***
```{r}
weatherData = read.csv("WeatherData/weather.csv")
weatherData
```

```{r}
rm(data)
dim(numberOfTrips)
data = merge(numberOfTrips, crimeCount, by.x = "date", by.y = "OCCURRED_ON_DATE")
dim(data)
cat("Total Number of na fields in the dataset", sum(is.na(data)))

weatherData$datetime = as.Date(weatherData$datetime)
data = merge(data, weatherData, by.x = "date", by.y ="datetime", all.x = TRUE)
dim(data)
cat("Total Number of na fields in the dataset", sum(is.na(data)))

sum(is.na(data$totalCrimes))
sum(is.na(data$temp))
sum(is.na(data$humidity))
sum(is.na(data$windspeed))
sum(is.na(data$visibility))
sum(is.na(data$feelslike))
sum(is.na(data$uvindex))


data
```

***Write the merged data into one big file. Now you can just get if from over here***
```{r}
write.csv(data, "MergedData/data.csv")
```

***Read CSV if got already written.***
```{r}
data = read.csv("MergedData/data.csv")
```

```{r}
library("ggplot2")
str(data)
ggp = ggplot(data = data) +
  geom_line(aes(x = date, y = totalTrips), color = 2)
ggp

ggplot(data = data) + 
  geom_line(aes(x = date, y = totalCrimes), color = 3) +
  geom_line(aes(x = date, y = temp), color = 4)
```
